---
title: "Examining COVID-19 data (web browser)"
output: 
  flexdashboard::flex_dashboard:
    vertical_layout: scroll
    orientation: rows
    source_code: embed
    theme: sandstone
runtime: shiny
---

```{r setup, include=FALSE}
# knitr opts
knitr::opts_chunk$set(cache = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      error = FALSE,
                      echo = FALSE,
                      eval = TRUE
)
```

```{r libraryload}
library(flexdashboard)
library(tidyverse)
library(lubridate)
library(tsibble)
library(jsonlite)
library(gridExtra)
# remotes::install_github("ropenscilabs/ochRe")
library(ochRe)
library(forcats)
library(plotly)
library(DT)
#library(googlesheets)
library(ggrepel)
library(shiny)
```

```{r read-jhu}
# JHU
covid_jh <- read_csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv")
covid_jh_oz_st <- covid_jh %>%
  filter(`Country/Region` == "Australia") %>%
  pivot_longer(cols=contains("/20"), names_to = "date") %>%
  mutate(date = mdy(date)) %>%
  rename(state = `Province/State`) %>%
  mutate(state = recode(state, "New South Wales"="NSW", "Victoria"="VIC", "Queensland"="QLD", 
                        "Australian Capital Territory"="ACT", "South Australia"="SA", 
                        "Western Australia"="WA", "Tasmania"="TAS", "Northern Territory"="NT",
                        "From Diamond Princess"="DP")) %>%
  rename(count = value) %>%
  select(state, date, count) 
# From https://en.wikipedia.org/wiki/List_of_Australian_states_and_territories_by_gross_state_product
state_pop <- tibble(state = c("NSW","VIC","QLD","SA","WA","TAS","NT","ACT","DP"), 
    pop = c(8089526, 6594804, 5095100, 1751693, 2621680, 426709, 245869, 420379, NA))
```

```{r read_more_jhu}
covid_jh_deaths <- read_csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv")
covid_jh_recovered <- read_csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_recovered_global.csv")
```

```{r monash_data}
# Data from covid-19-au project
state_monash <- fromJSON("https://raw.githubusercontent.com/covid-19-au/covid-19-au.github.io/prod/src/data/state.json")
x <- tibble(date=state_monash)
x_wide <- x %>% unnest_wider(date) 
x_wide_bef19 <- x_wide[1:18,] %>% 
  pivot_longer(cols = NSW:NT, names_to = "state", values_to = "count") %>%
  unnest_wider(count) %>%
  rename(count=`...1`) %>%
  mutate(date = rep(names(state_monash)[1:18], rep(8, 18)),
         deaths = NA, recovered = NA, tested = NA) %>%
  select(date, state, count, deaths, recovered, tested)
x_wide_aft18 <- x_wide[19:length(state_monash),] %>% 
  pivot_longer(cols = NSW:NT) %>% 
  unnest_wider(value) %>%
  rename(state=name, 
         count=`...1`,
         deaths=`...2`,
         recovered=`...3`,
         tested=`...4`) %>%
  mutate(date = rep(names(state_monash)[19:length(state_monash)], 
                    rep(8, length(state_monash)-18))) %>%
  select(date, state, count, deaths, recovered, tested)
state_monash_data <- bind_rows(x_wide_bef19, x_wide_aft18) 
state_monash_data <- left_join(state_monash_data, state_pop)
state_monash_data <- state_monash_data %>%
  mutate(rate = round(count/pop * 100000, 2),
         test_rate = round(tested/pop * 1000), 2) %>%
  mutate(date = ymd(date))
#locations_monash <- fromJSON("https://raw.githubusercontent.com/covid-19-au/covid-19-au.github.io/prod/src/data/mapdata.json")
#test_loc_monash <- frmJSON("https://raw.githubusercontent.com/covid-19-au/covid-19-au.github.io/prod/src/data/mapdataHos.json")
```

Australia
=======================================================================

Row {data-height=350}
-----------------------------------------------------------------------

### *Latest incidence (per 100k) and testing rates (per 1000). States ordered by incidence rate*

<!-- # Now make plots using Monash COVID-19 data -->

```{r monash_test_rate, fig.width=10, fig.height=5, out.width="100%"}
ord <- state_monash_data %>% 
  filter(date == max(date)) %>%
  arrange(rate) %>%
  select(state)
covid_ne_jhu_smry <- state_monash_data %>% 
  filter(date == max(date)) %>% 
  mutate(state = factor(state, levels=ord$state))
p1 <- covid_ne_jhu_smry %>%
  ggplot(aes(x=state, y=rate, fill=state)) +
  geom_col() +
  scale_fill_ochre(palette = "mccrea") +
  ggtitle("Incidence") + 
  coord_flip() + xlab("") + ylab("Rate") +
  theme(legend.position = "none")
p2 <- covid_ne_jhu_smry %>%
  ggplot(aes(x=state, y=test_rate, fill=state)) +
  geom_col() +
  scale_fill_ochre(palette = "mccrea") +
  ggtitle("Testing") + 
  coord_flip() + xlab("") + ylab("Rate") +
  theme(legend.position = "none")
grid.arrange(p1, p2, ncol=2)
```

### Data

```{r monash_table_counts}
covid_ne_jhu_smry %>% 
  select(state, date, count, pop, tested, rate, test_rate, deaths) %>%
  arrange(desc(count)) %>%
  datatable(
    rownames = FALSE,
    options = list(
      bPaginate = FALSE,
      pageLength = 4
    ))
```

Row {data-height=350}
-----------------------------------------------------------------------

###  COVID-19 rate per 100k people by state in Australia

```{r monash_state_incidence, fig.width=10,  fig.height=5.5}
renderPlot({
  state_monash_data %>% 
  mutate(state = factor(state, levels=ord$state[9:1])) %>%
    ggplot(aes(x=date, y=rate)) +
      geom_point(aes(colour=state)) +
      geom_smooth(se=FALSE, colour="black") + 
      scale_colour_ochre("", palette = "mccrea") +
      facet_wrap(~state, ncol = 4, nrow = 2) + 
      xlab("") + ylab("Rate (per 100k)") +
    guides(colour = guide_legend(nrow = 2, byrow = TRUE)) +
    theme(legend.position = "right",
          legend.direction = "horizontal",
          legend.title = element_blank())
})
```

Melbourne
=======================================================================

Inputs {.sidebar data-width=200}
-----------------------------------------------------------------------

Pedestrian traffic in Melbourne in the last week in comparison to a year ago. 



```{r eval=FALSE}
library(rwalkr)
#pedestrian <- melb_walk_fast(year=c(2019, 2020), sensor="Southern Cross Station") 

ped_2020 <- melb_walk(from=Sys.Date() - 7L)
ped_2019 <- melb_walk(from=Sys.Date() - 30L - years(1), to=Sys.Date() - years(1))


ped_2019 <- ped_2019 %>%
  filter(Date != today()) %>%
  mutate(wk =  week(Date), 
         wday = wday(Date, label = TRUE),
         year = year(Date))
ped_2020 <- ped_2020 %>%
  mutate(wk =  week(Date), 
         wday = wday(Date, label = TRUE),
         year = year(Date))

ped <- bind_rows(ped_2019, ped_2020) %>%
  mutate(wday = factor(wday, 
                       levels = wday(max(ped_2020$Date)-days(6:0), label=TRUE)))
save(ped, file="data/ped.rda")
```

```{r}
load("data/ped.rda")
period <- ped %>% filter(year == 2020) %>% select(Date) %>% summarise(mn = min(Date), mx = max(Date))
selectInput("sensor", "Choose a sensor:", 
              choices = unique(ped$Sensor), 
              selected = "Melbourne Central",  
              width="100%")
```

Row {data-height=400}
-----------------------------------------------------------------------

```{r out.width="100%", fig.width=10, fig.height=4}
renderPlot({
  ped %>%
  filter(Sensor == input$sensor) %>%
  ggplot(aes(x=Time, y=Count, group=Date, colour=as.factor(year))) +
    geom_line() +
    facet_wrap(~wday, ncol=7) + 
    scale_colour_ochre("year", palette="emu_woman_paired") + 
  theme(legend.position="bottom",
          legend.title = element_blank()) +
  ggtitle(paste(period$mn, "to", period$mx))
})
```

World Incidence
=======================================================================


Inputs {.sidebar data-width=200}
-----------------------------------------------------------------------

```{r choose_countries}
covid_jh_long <- covid_jh %>%
  pivot_longer(cols=contains("/20"), names_to = "date") %>%
  mutate(date = mdy(date)) %>%
  group_by(date, `Country/Region`) %>%
  summarise(count = sum(value)) %>%
  ungroup() %>%
  group_by(`Country/Region`) %>%
  mutate(daily = count-dplyr::lag(count, order_by = date)) %>%
  mutate(new_past_7d = count - dplyr::lag(count, 7, order_by = date)) %>%
  ungroup() %>%
  arrange(`Country/Region`) %>%
  filter(`Country/Region` != "Diamond Princess")
covid_since100 <- covid_jh_long %>%
  filter(count > 100) %>%
  group_by(`Country/Region`) %>%
  mutate(days = as.numeric(date - min(date))) %>%
  ungroup() 
keep <- covid_since100 %>%
  count(`Country/Region`) %>%
  filter(n>14)
covid_since100 <- covid_since100 %>%
  filter(`Country/Region` %in% keep$`Country/Region`)
covid_since100_labels <- covid_since100 %>%
  group_by(`Country/Region`) %>%
  filter(days == max(days)) %>%
  ungroup()
checkboxGroupInput("cnt", "Choose countries:", choices = unique(covid_jh_long$`Country/Region`), selected = c("Australia", "Germany", "US", "Sweden", "Italy", "Spain", "Korea, South"),  width="100%")
```


Row {data-height=400}
-----------------------------------------------------------------------

```{r incidence_rates, out.height="100%", fig.width=10, fig.height=10}
renderPlot({
china <- filter(covid_since100, `Country/Region`=="China")
china_label <- filter(covid_since100_labels, `Country/Region`=="China")
countries <- covid_since100 %>%
  filter(`Country/Region` %in% input$cnt)
countries_labels <- filter(covid_since100_labels, `Country/Region` %in% input$cnt)
p3 <- 
  ggplot() +
  geom_line(data=china, aes(x=days, y=count), colour="black") +
  geom_text_repel(data=china_label, aes(x=days, y=count, label=`Country/Region`), colour="black") +
  geom_line(data=countries, aes(x=days, y=count, colour=`Country/Region`)) + 
  geom_text_repel(data=countries_labels, aes(x=days, y=count,
       label=`Country/Region`, colour=`Country/Region`)) + 
  scale_y_log10() + 
  xlab("Days since 100 cases") + 
  ylab("Incidences (log scale)") +
  scale_colour_brewer("", palette="Dark2") +
  theme(legend.position = "none") 
china_trend <- covid_jh_long %>% 
    filter(`Country/Region` == "China")
china_trend_labels <- china_trend %>%
  filter(date == max(date))
countries_trend  <- covid_jh_long %>% 
    filter(`Country/Region` %in% input$cnt) 
countries_trend_labels <- countries_trend %>%
  group_by(`Country/Region`) %>%
  filter(date == max(date)) %>%
  ungroup()
p4 <- ggplot() + 
      geom_line(data=countries_trend, aes(x=count, y=new_past_7d,
                                     colour=`Country/Region`)) + 
      geom_text_repel(data=countries_trend_labels, aes(x=count, y=new_past_7d,
       label=`Country/Region`, colour=`Country/Region`)) + 
      geom_line(data=china_trend, aes(x=count, y=new_past_7d), colour="black") +
        xlab("Total cases") + ylab("Cases in past 7 days") +
      geom_text_repel(data=china_trend_labels, aes(x=count, y=new_past_7d,
       label=`Country/Region`), colour="black") + 
      scale_colour_brewer("", palette="Dark2") +
      theme(legend.position = "none") 
#p4 <- covid_jh_long %>% 
#    filter(`Country/Region` %in% input$cnt) %>%
#    ggplot(aes(x=date, y=count)) + geom_col() +
#    facet_wrap(~`Country/Region`, scales="free_y") + 
#    scale_y_log10() + xlab("") + ylab("Count (log scale)")
grid.arrange(p3, p4, ncol=2)
})
```

Row {data-height=400}
-----------------------------------------------------------------------

### Incidence by country `r max(covid_jh_long$date)`

```{r country_counts}
covid_jh_long %>% 
  filter(date == max(date)) %>%
  select(`Country/Region`, count, daily) %>%
  arrange(`Country/Region`) %>%
  datatable(options = list(
  bPaginate = FALSE
))
```

NOT Mortality rates {data-orientation=columns}
=======================================================================

Column {.sidebar data-width=150}
-----------------------------------------------------------------------
The rates are calculated by taking 1-deaths/incidence. Only countries with more than 500 cases are shown.

<p> *Its more usual to talk about mortality rates. From a personal perspective, I'd rather hear about proportion of people who are still fighting it or recovered, than mortality.*

Row {data-height=150}
-----------------------------------------------------------------------

### Counts


```{r survival}
covid_jh_deaths_long <- covid_jh_deaths %>%
  pivot_longer(cols=contains("/20"), names_to = "date") %>%
  mutate(date = mdy(date)) %>%
  group_by(date, `Country/Region`) %>%
  summarise(deaths = sum(value)) %>%
  ungroup()
covid_jh_all <- left_join(covid_jh_long, covid_jh_deaths_long)
latest_surv <- covid_jh_all %>%
  select(`Country/Region`, date, count, deaths) %>%
  filter(date == max(date)) %>%
  mutate(
    notmortality = round(1 - deaths/count, 4),
    mortality = round(deaths/count, 4)) %>%
  filter(!is.na(notmortality)) %>%
  filter(count > 500)
latest_surv <- latest_surv %>%
  mutate(`Country/Region` = fct_reorder(`Country/Region`, notmortality))
```

```{r plot-counts, fig.height=7, fig.width=3, out.width="50%"}
psurv1 <- ggplot(data=latest_surv, aes(x=`Country/Region`, y=count)) +
  geom_col(aes(text=paste0("not mortality: ", notmortality))) + 
  #scale_y_log10() +
  coord_flip() + xlab("") + ylab("Count") # (log scale)") 
renderPlotly({
  ggplotly(psurv1)
})
```

Row {data-height=150}
-----------------------------------------------------------------------

### Not mortality

```{r plot_survival, fig.height=7, fig.width=3, out.width="50%"}
latest_surv_long <- latest_surv %>%
  pivot_longer(cols = -c(`Country/Region`, date, count, deaths), names_to = "type", values_to = "rate") %>%
  mutate(type = factor(type, levels = c("notmortality", "mortality")))
psurv2 <- ggplot(data=latest_surv_long, aes(x=`Country/Region`, y=rate, fill =type, text=paste0("count: ", count)),  stat="identity") +
  geom_bar(position="fill", stat="identity") +
  scale_fill_brewer("", palette="Dark2") +
  coord_flip() + xlab("") + ylab("Rate") +
  theme(legend.position="none")
  #geom_hline(yintercept=1, colour = "red") +
  #ylim(c(0,1))
renderPlotly({
  ggplotly(psurv2)
})
```

Recovery
=======================================================================

Row {.sidebar data-width=350}
-----------------------------------------------------------------------

**Exploring the incidence and recovery timeline for China, and Australia.**

Daily incidence is plotted, along with daily recovered counts. The form looks like many of the epidemic simulation data, for example https://alhill.shinyapps.io/COVID19seir/. 

Row {data-width=650}
-----------------------------------------------------------------------

```{r}
library(tidyquant)
# Examine daily counts and recovery for countries with early outbreaks, eg China
china_count <- covid_jh %>% 
  filter(`Country/Region` == "China") %>%
  pivot_longer(cols=contains("/20"), names_to = "date") %>%
  mutate(date = mdy(date)) %>%
  group_by(date) %>%
  summarise(count = sum(value)) %>%
  ungroup() %>%
  mutate(daily = count - dplyr::lag(count, order_by = date))
china_recovered <- covid_jh_recovered %>% 
  filter(`Country/Region` == "China") %>%
  pivot_longer(cols=contains("/20"), names_to = "date") %>%
  mutate(date = mdy(date)) %>%
  group_by(date) %>%
  summarise(count = sum(value)) %>%
  ungroup() %>%
  mutate(recovered = count - dplyr::lag(count, order_by = date))
china <- left_join(china_count, china_recovered, by="date") %>%
  select(-count.x, -count.y) %>%
  pivot_longer(cols = c("daily", "recovered"), names_to = "type", values_to = "count") %>%
  mutate(type = factor(type, levels = c("recovered", "daily")))
p_china <- ggplot(china, aes(x=date, y=count, colour=type)) + 
  geom_hline(yintercept=0, colour="black") +
  geom_point(alpha=0.5) + geom_ma(n=7, linetype=1, size=2) + 
  scale_colour_brewer("", palette = "Dark2") +
  annotate("text", x=china$date[1], y=14500, label="China", hjust=0) +
  ylim(c(0, 7500)) +
  xlab("") + ylab("Daily counts") +
  theme(legend.position = "none")
```

```{r}
# Examine daily counts and recovery for countries with early outbreaks, eg China
australia_count <- covid_jh %>% 
  filter(`Country/Region` == "Australia") %>%
  pivot_longer(cols=contains("/20"), names_to = "date") %>%
  mutate(date = mdy(date)) %>%
  group_by(date) %>%
  summarise(count = sum(value)) %>%
  ungroup() %>%
  mutate(daily = count - dplyr::lag(count, order_by = date)) 
australia_recovered <- covid_jh_recovered %>% 
  filter(`Country/Region` == "Australia") %>%
  pivot_longer(cols=contains("/20"), names_to = "date") %>%
  mutate(date = mdy(date)) %>%
  group_by(date) %>%
  summarise(count = sum(value)) %>%
  ungroup() %>%
  mutate(recovered = count - dplyr::lag(count, order_by = date))
australia <- left_join(australia_count, australia_recovered, by="date") %>%
  select(-count.x, -count.y) %>%
  pivot_longer(cols = c("daily", "recovered"), names_to = "type", values_to = "count") %>%
  mutate(type = factor(type, levels = c("recovered", "daily")))
#australia <- australia %>% 
#  as_tsibble(index=date, key=type) %>%
#  mutate(ma7 = slide_dbl(count, mean, .size=7))
p_australia <-  ggplot(australia, aes(x=date, y=count, colour=type)) + 
  geom_hline(yintercept=0, colour="black") +
  geom_point(alpha=0.5) + geom_ma(n=7, linetype=1, size=2) +
  scale_colour_brewer("", palette = "Dark2") +
  annotate("text", x=australia$date[1], y=450, label="Australia", hjust=0) +
  xlab("") + ylab("Daily counts") +
  theme(legend.position = "bottom")
```

```{r}
library(patchwork)
renderPlot(
  height = 600,
  {p_china/p_australia}
)
#grid.arrange(p_china, p_australia, ncol=1)
```

Sources
=======================================================================


Data is taken from several sources. 

The incidence data is from John Hopkins University,  https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_time_series. 

Population for Australian states is from https://en.wikipedia.org/wiki/List_of_Australian_states_and_territories_by_gross_state_product. 

Australian data is sync'd with "The Real-time COVID-19 Status in Australia" https://github.com/covid-19-au/covid-19-au.github.io. This data is sourced from media reports and is curated by a team at Monash University. 

Earlier versions also used data collated by Nick Evershed on Australian records at https://interactive.guim.co.uk/docsdata/1q5gdePANXci8enuiS4oHUJxcxC13d6bjMRSicakychE.json. 

Earlier version extracted testing data for Australia from https://ourworldindata.org/covid-testing. And NSW is updated from https://www.health.nsw.gov.au/news/Pages/20200323_00.aspx. Both of these sites need some manual curation, because one needs a manual download, and the second needs the url updated with the current date, and a check that tables are still the same.

All code is available at https://github.com/dicook/covid19. Suggestions and pull requests encouraged. 

